Numerical modelling of the Cordilleran ice sheet

### 2014-07-20 kappa.tex
Remove comment replies.

% My commands
\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
\newcommand{\todo}[1]{} %\emph{[\textbf{Todo:} #1]}}
\newcommand{\mref}[0]{\textbf{[ref.]}}

@@

\begin{description}
  \item[\CCLI] \bibentry{Seguinot.etal.2014}.
  \item[\PSDV] \bibentry{Seguinot.2013}.
  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}.
  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.\renote{
    Arjen, you had several comments regarding the style of references
    throughout the kappa (name formatting, italic text). I am using the
    Copernicus style here, but is there another recommendation to follow?
    Maybe Peter knows?}
\end{description}

@@

In our daily life, we most commonly experience ice as small, solid cubes
that crackle on contact with refreshing liquids. Yet, on geological
time scales, and under high pressure, ice behaves quite differently.
In the cold and mountainous regions of the globe, ice accumulates through
cumulative snowfall until it reaches a critical thickness. Beyond this
threshold, usually in excess of 50\,m, but dependent on temperature, ice
deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
\emph{Glaciers}, these bodies of moving ice, move by a combination of
meltwater-induced sliding at the base \citep[\S532]{Saussure.1786}\renote{
    Actually I mean volume 1, published in ``MDCCLXXXVI'' (1786, right?),
    which as a couple of beautiful paragraphs (would be worth
    translating if I had time) about how the "heat from the Earth's interior"
    melts basal ice to lubricate glacier flow. This is precisely those
    paragraphs that \citet{Forbes.1846} refers to sixty years later when
    advancing his ``viscous theory of glacier notion''.},
and viscous deformation within the
ice body \citep{Forbes.1846b}.

As glaciers flow and slide across their bed, they transport\renote{
    I would like to keep this statement general, surface transport is also
    important. Erosion is not only by basal rock debris but also, indirectly,
    by glacial meltwater.}
rock debris and erode the landscape, thereby leaving geomorphological traces of
their former presence (\cref{fig:glaciation-indicators}). In several parts of
the world, montane people\renote{
    I insist on keeping this: people of the Alps, for istance, had understood
    that glacial erratics originated from glaciers before it has ever been
    documented by explorers and scientists. It is stated in the old literature.
    Check the new reference, reprinted in \citet{Mathews.1898} (not quite sure
    how to cite this material).}
and early explorers learned since long to read these traces, and to
understand that glaciers had once been more extensive than today
\citep[e.g.,][p.~21]{Windham.Martel.1744}. With the emergence of the glacial
theory came the idea that, under colder temperatures
than today, expansive \emph{ice sheets} had once covered much of Europe and
North America \citep{Agassiz.1840}\renote{
    I am a bit reluctant to write ``\citet{Agassiz.1840} eventually posed the
    glacial theory''. Although he published these ideas, do they really
    originate from him?}.

@@

Nevertheless, the landform record is sparse in time and, most often, spatially
incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
lanscape, and much of this evidence has been overprinted by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994,
Kleman.etal.2006, Kleman.etal.2010}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical ice sheet models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
help. Although these models alone contain too many uncertainties to reproduce
accurate ice sheet reconstructions during former glacial cycles
\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
data where they exist \citep{Napieralski.etal.2007}, while allowing
physics-based, quantitative reconstructions of ice sheet sectors where they do
not
\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
In addition, numerical ice sheet models provide quantitative insights on the
four-dimensional distribution of stresses and deformations in palaeo-ice sheets
and allow for reconstructions of their past thickness and volume evolution, which
are difficult to infer from geological evidence\renote{
    Is this the place where to cite \citep{Denton.Hughes.1981}? \todo{I need to
    check the contents of this book (library).}
}.

@@

During the last glacial cycle, much of North America has been covered by ice
(\cref{fig:map-northamerica}). During the Last Glacial Maximum (LGM), peak of
glaciation, three major ice sheets had collided to
form a continuous blanket of ice that spanned from southern Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago\renote{
    Although already a few 100\,km into the ice margin, I think that Chicago
    is a more illustrative location than Ohio or smaller cities there.}.
The Laurentide ice sheet, largest of the three, was centred on the Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the \emph{Cordilleran ice sheet}, intermediate in size,
capped the mountain
ranges of the Western Cordillera.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effects on
vegetation are clearly expressed in the landscape
(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
region, mean annual \emph{temperatures} vary significantly from north to south
(\cref{fig:plot-atm}a). About half of the model domain has mean annual
temperatures below freezing. However, temperature \emph{seasonality}\renote{
    ``Seasonality'' is a common term in the meteorological literature,
    implicitly referring to temperature. At least this is what I understood
    from Qiong.},
here computed as the difference between the warmest and the coldest months,
varies between around 10\,\degC\ along most of the Pacific Coast to above
40\,\degC\ for the interior Arctic sectors\renote{
    I think the word ``polar'' or ``Arctic'' is needed here, see
    \cref{fig:plot-atm}b.}
(\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperatures are well below
freezing point experience warm summers.

@@

The retreat of the ice sheet during the last \emph{deglaciation} was likely
rapid. Most of the evidence for the style and pace of deglaciation comes from
the Fraser Plateau. Studies there have indicated that some of the highest
uplands emerged from the ice sheet prior to the surrounding lowlands, possibly
resulting in thin, stagnant ice remnants covering parts of the plateau
\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
is known about the deglaciation of the northern sector of the Cordilleran ice
sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of deglaciation, with active retreat of the ice margin towards
the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in
\CCYC\renote{
    I have not had time to dig into new literature. The Ward paper on
    deglacial valley fills in the Pelly River is like Chinese to me, and I
    could not access the Dyke paper on the Liard Lowland, assuming this is the
    one cited by Martin in his paper.}.

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$\renote{
  $\grad$, $\div$, $\tgrad$, $\tdiv$, are the classic gradient and divergence operators, applied on vector and tensor fields. $\tr$ is the trace of a matrix. I don't think these need to be defined.},
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
values measured below and above a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

@@

The till cohesion $c_0$ is assumed to be zero\renote{
  Yes, we should probably defend this choice, but I am not aware of any
  observations justifying this.}.
The effective pressure on the
till, $N$, is determined by the modelled amount of water at the bed. However,
two different parametrisations are used within this thesis. In {\CCLI}, using
PISM~0.5, the effective pressure is linearly related to water content by a
simple parametrisation \citep[Eq.~13]{Winkelmann.etal.2011},
\begin{equation}
    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,,
\end{equation}
where $W$ is the amount of water at the bed, and varies from zero to
$W_{max}=2$\,m, a threshold above which further melt simply exits the system,
thus assuming instantaneous drainage. Although mass-conserving water drainage
is now implemented in PISM \citep{Bueler.Pelt.2014}, it is not included in the
simulations presented here for the sake of simplicity.

@@

At the interface between the ice sheet and the atmosphere
(\cref{fig:model-interfaces}),
surface mass fluxes are computed from monthly mean surface air temperature,
$T_m$, and monthly precipitation, $P_m$, using a temperature-index model
\citep[\cref{fig:model-pdd}; cf.][]{Hock.2003}. Snow accumulation, $A_s$,
equals precipitation
when temperature is below ${T_0=0\,\degC}$, and decreases to zero\renote{
    Well spotted, thanks!}
linearly with temperature between $T_0$ and ${T_1=2\,\degC}$,
\begin{equation}
    A_s =
    \begin{cases}
        P_m     & \text{if}\ T_m \le T_0 \,, \\
        P_m \frac{T_1-T_m}{T_1-T_0}
                & \text{if}\ T_0 < T_m < T_1 \,, \\
        0       & \text{if}\ T_1 \le T_m \,,
    \end{cases}
\end{equation}

>>>>

% My commands
%\newcommand{\renote}[1]{\footnote{\textbf{Comment reply}: #1}}
\newcommand{\todo}[1]{} %\emph{[\textbf{Todo:} #1]}}
%\newcommand{\mref}[0]{\textbf{[ref.]}}

@@

\begin{description}
  \item[\CCLI] \bibentry{Seguinot.etal.2014}.
  \item[\PSDV] \bibentry{Seguinot.2013}.
  \item[\PSDP] \bibentry{Seguinot.Rogozhina.2014}.
  \item[\CCYC] \bibentry{Seguinot.etal.manuscript}.
\end{description}

@@

In our daily life, we most commonly experience ice as small, solid cubes
that crackle on contact with refreshing liquids. Yet, on geological
time scales, and under high pressure, ice behaves quite differently.
In the cold and mountainous regions of the globe, ice accumulates through
cumulative snowfall until it reaches a critical thickness. Beyond this
threshold, usually in excess of 50\,m, but dependent on temperature, ice
deforms under its own weight to flow downhill (\cref{fig:glacier-mechanics}).
\emph{Glaciers}, these bodies of moving ice, move by a combination of
meltwater-induced sliding at the base \citep[\S532]{Saussure.1786},
and viscous deformation within the
ice body \citep{Forbes.1846b}.

As glaciers flow and slide across their bed, they transport
rock debris and erode the landscape, thereby leaving geomorphological traces of
their former presence (\cref{fig:glaciation-indicators}). In several parts of
the world, montane people
and early explorers learned since long to read these traces, and to
understand that glaciers had once been more extensive than today
\citep[e.g.,][p.~21]{Windham.Martel.1744}. With the emergence of the glacial
theory came the idea that, under colder temperatures
than today, expansive \emph{ice sheets} had once covered much of Europe and
North America \citep{Agassiz.1840}.

@@

Nevertheless, the landform record is sparse in time and, most often, spatially
incomplete. Palaeo-ice sheets did not leave a continuous imprint on the
lanscape, and much of this evidence has been overprinted by subsequent glacier
re-advances and other geomorphological processes \citep{Kleman.1994,
Kleman.etal.2006, Kleman.etal.2010}.
Therefore, geomorphological reconstructions of former glacier extent require a
substantial amount of interpretation. This is where numerical ice sheet models,
incorporating the processes of glacier sliding \citep[e.g.,][]{Weertman.1957}
and deformation \citep[e.g.,][]{Nye.1953} observed on modern glaciers, can
help. Although these models alone contain too many uncertainties to reproduce
accurate ice sheet reconstructions during former glacial cycles
\citep[e.g.,][]{Hebeler.etal.2008}, they can be calibrated against geological
data where they exist \citep{Napieralski.etal.2007}, while allowing
physics-based, quantitative reconstructions of ice sheet sectors where they do
not
\citep[e.g.,][]{Marshall.etal.2002, Tarasov.Peltier.2004}.
In addition, numerical ice sheet models provide quantitative insights on the
four-dimensional distribution of stresses and deformations in palaeo-ice sheets
and allow for reconstructions of their past thickness and volume evolution, which
are difficult to infer from geological evidence.

@@

During the last glacial cycle, much of North America has been covered by ice
(\cref{fig:map-northamerica}). During the Last Glacial Maximum (LGM), peak of
glaciation, three major ice sheets had collided to
form a continuous blanket of ice that spanned from southern Alaska to
Newfoundland, connecting with the Greenland ice sheet in the Canadian Arctic,
and extending as far south as Chicago.
The Laurentide ice sheet, largest of the three, was centred on the Hudson Bay,
and fully covered the Canadian Prairies, Eastern Canada, and the Great Lakes
Basin. The Innuitian ice sheet, much smaller, covered the northernmost Canadian
archipelago. Finally, the \emph{Cordilleran ice sheet}, intermediate in size,
capped the mountain
ranges of the Western Cordillera.

@@

The topographic complexity of the North American Cordillera is associated with
distinctive patterns of temperature and precipitation, whose effects on
vegetation are clearly expressed in the landscape
(\cref{fig:photo-fraser-valley}). Due to the large latitudinal extent of the
region, mean annual \emph{temperatures} vary significantly from north to south
(\cref{fig:plot-atm}a). About half of the model domain has mean annual
temperatures below freezing. However, temperature \emph{seasonality},
here computed as the difference between the warmest and the coldest months,
varies between around 10\,\degC\ along most of the Pacific Coast to above
40\,\degC\ for the interior Arctic sectors
(\cref{fig:plot-atm}b). There exists a
partial correlation between cold regions and those that have a large
seasonality, so that many regions where mean annual temperatures are well below
freezing point experience warm summers.

@@

The retreat of the ice sheet during the last \emph{deglaciation} was likely
rapid. Most of the evidence for the style and pace of deglaciation comes from
the Fraser Plateau. Studies there have indicated that some of the highest
uplands emerged from the ice sheet prior to the surrounding lowlands, possibly
resulting in thin, stagnant ice remnants covering parts of the plateau
\citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}. Less
is known about the deglaciation of the northern sector of the Cordilleran ice
sheet, but abundant glacial landforms from the Liard Lowland indicate a
different mode of deglaciation, with active retreat of the ice margin towards
the surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in
\CCYC.

@@

Thirdly, after defining the strain-rate tensor
${\SRT = \frac{1}{2}(\tgrad{\vv} + \tgrad{\vv}^{\mathrm{T}})}$,
and the deviatoric stress tensor, ${\DST = \CST + p\,\IDT}$,
where $p=-\frac{1}{3}\tr(\CST)$ is the hydrostatic pressure and
$\tens{\delta}$ is the identity tensor, a \emph{constitutive law} for ice
\citep{Nye.1953} based on laboratory experiments \citep{Glen.1952} is used to
relate these two quantities. It reads:
\begin{equation}
    \label{eqn:glenslaw}
    \SRT = A\,\tau_e^{n-1}\,\DST \,.
\end{equation}
The equivalent stress, $\tau_e$, is defined by
${\tau_e}^2 = -\mathrm{II}_{\DST} = \frac{1}{2} \tr(\DST^2)$,
where $\mathrm{II}_{\DST}$ is the second invariant of the stress tensor.
The ice softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$, and
water content, $\omega$, through an Arrhenius-type law
\citep[Eqs.~63--65]{Paterson.Budd.1982, Aschwanden.etal.2012},
\begin{equation}
    A =
    \begin{cases}
        A_c (1+f\omega)\,e^\frac{-Q_c}{RT_{pa}}
            & \text{if}\ T < T_c \,, \\
        A_w (1+f\omega)\,e^\frac{-Q_w}{RT_{pa}}
            & \text{if}\ T \ge T_c \,,
    \end{cases}
\end{equation}
where $T_{pa}$ is the pressure-adjusted temperature calculated using the
Clapeyron relation, ${T_{pa} = T - \beta p}$. $R$ is the ideal gas constant,
and $A_c$, $A_w$, $Q_c$ and $Q_w$, are constant parameter corresponding to
values measured below and above a critical temperature threshold
${T_c=-10}$\,\degC\ \citep[\cref{tab:iceflowparams};][]{Paterson.Budd.1982}.
The water fraction, $\omega$, is capped at a maximum value of 0.01, above which
no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
Greve.1997}.

@@

The till cohesion $c_0$ is assumed to be zero.
The effective pressure on the
till, $N$, is determined by the modelled amount of water at the bed. However,
two different parametrisations are used within this thesis. In {\CCLI}, using
PISM~0.5, the effective pressure is linearly related to water content by a
simple parametrisation \citep[Eq.~13]{Winkelmann.etal.2011},
\begin{equation}
    N = \rho gh (1 - \alpha \frac{W}{W_{max}}) \,,
\end{equation}
where $W$ is the amount of water at the bed, and varies from zero to
$W_{max}=2$\,m, a threshold above which further melt simply exits the system,
thus assuming instantaneous drainage. Although mass-conserving water drainage
is now implemented in PISM \citep{Bueler.Pelt.2014}, it is not included in the
simulations presented here for the sake of simplicity.

@@

At the interface between the ice sheet and the atmosphere
(\cref{fig:model-interfaces}),
surface mass fluxes are computed from monthly mean surface air temperature,
$T_m$, and monthly precipitation, $P_m$, using a temperature-index model
\citep[\cref{fig:model-pdd}; cf.][]{Hock.2003}. Snow accumulation, $A_s$,
equals precipitation
when temperature is below ${T_0=0\,\degC}$, and decreases to zero
linearly with temperature between $T_0$ and ${T_1=2\,\degC}$,
\begin{equation}
    A_s =
    \begin{cases}
        P_m     & \text{if}\ T_m \le T_0 \,, \\
        P_m \frac{T_1-T_m}{T_1-T_0}
                & \text{if}\ T_0 < T_m < T_1 \,, \\
        0       & \text{if}\ T_1 \le T_m \,,
    \end{cases}
\end{equation}
